---
title: "Final Project: Walkability and Public Health in the US"
author: "Lou Godmer, Kienan Battin, Divakar Mehta"
date: "April 17, 2023"
output:
    pdf_document: default
    toc: true
    number_sections: true
    toc_depth: 4
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\tableofcontents 
\newpage


## Objective

The objective is to quantify the causal effect that the "walkability" of a region has on the public health indicators: obesity and physical activity. 
The original data comes from two sources:
1. The [U.S. Chronic Disease Indicators](https://chronicdata.cdc.gov/Chronic-Disease-Indicators/U-S-Chronic-Disease-Indicators-CDI-/g4ie-h725) provides reported cases of a set of 124 indicators that are important to public health, and the geographic location of the case. 
2. The [Walkability Index](https://catalog.data.gov/dataset/walkability-index) quantifies every Census 2019 block group's relative "Walkability" as defined by the EPA based on characteristics such as easy walking access to public transit, jobs, stores and services. 
Quantifying the causal effect of walkability on public health can help policy makers understand how community planning measures that may improve or degrade the walkability of the region will impact public health.

The appendix of this document describes the pre-processing methodology that was used combine the two data sets to enable the quantitative analysis. Because the pre-processing methodology can take an hour or more to execute, we exported the pre-processed data and made it available for download in a publicly accessible location. The beginning of this document imports the pre-processed data and the rest of the analysis is done based on the pre-processed data.

## Load necessariy libararies

```{r load libraries, message=FALSE, warning=FALSE}

rm(list=ls())

options(repos = list(CRAN="http://cran.rstudio.com/"))

if (!require('NHANES')) install.packages('NHANES')
library('openxlsx')

if (!require('ggplot2')) install.packages('ggplot2')
library('ggplot2')

if (!require('dplyr')) install.packages('dplyr')
library('dplyr')

if (!require('GGally')) install.packages('GGally')
library('GGally')

if (!require('tableone')) install.packages('tableone')
library(tableone) 

if (!require('pROC')) install.packages('pROC')
library(pROC)

if (!require('tidyr')) install.packages('tidyr')
library(tidyr)

if (!require('tidycensus')) install.packages('tidycensus')
library(tidycensus)

if (!require('tigris')) install.packages('tigris')
library(tigris)

if (!require('sf')) install.packages('sf')
library(sf)

if (!require('stringr')) install.packages('stringr')
library(stringr)

if (!require('dplyr')) install.packages('dplyr')
library(dplyr)

```


## Load the data

Download the data which has already undergone the pre-processing methodology described in the appendix.
WARNING: this may take several minutes. To avoid unnecessary downloads, the commands are commented out. Un-comment and execute the commands to download the data.

```{r Download the data, echo=TRUE }
#download.file("https://walkabilityandhealth.blob.core.windows.net/walkabilityandhealth/disease_with_walkability.zip", destfile="disease_with_walkability.zip")
#unzip("disease_with_walkability.zip", "disease_with_walkability.csv")
```
```{r load the data}

disease_with_walkability <- read.csv("disease_with_walkability.csv")

```

## Shape the data set for our analysis purposes

The walkability data has observations with information about a BlockGroup. BlockGroup is a geographic unit used by the US Census Bureau and is a subdivision of a census tract, which in turn is a subdivision of a county. Block groups usually contain between 600 and 3000 people and are the smallest standardized unit of geography for which the Walkability score is measured. All of our analysis will be based on block group as the observation unit. 

The disease indicator data has a column for "Question" which contains the details of the disease indicator that was measured. Of the 203 total questions we have narrowed down to 12 that will be used in our analysis. Questions were selected which have the same unit of measure - Crude Prevalence - which means that the value is the percentage of the overall population. For simplicity we have also selected the more general questions rather than inter-sectional questions. For example we chose "Arthritis among adults aged >= 18 years" rather than "Arthritis among adults aged >= 18 years who have heart disease". The result of the selection process, along with how we intend to use each variable in our analysis, is in the table below:

| Topic | Question | QuestionID | Intended Usage |
| ---------- | -------------- | -------- | -------- | -------- |
| Nutrition, Physical Activity and Weight Status | Overweight or obesity among adults aged >= 18 years | NPAW2_1 | Dependent Var |
| Nutrition, Physical Activity and Weight Status | Meeting aerobic physical activity guidelines for substantial health benefits among adults aged >= 18 years | NPAW11_1 | Dependent var |
| Alcohol | Heavy drinking among adults aged >= 18 years | ALC5_1 | Independent Var |
| Arthritis | Arthritis among adults aged >= 18 years | ART1_1 | Independent var |
| Asthma | Current asthma prevalence among adults aged >= 18 years | AST1_1 | Independent var |
| Chronic Obstructive Pulmonary Disease | Prevalence of chronic obstructive pulmonary disease among adults >= 18 | COPD2_0 | Independent var |
| Oral Health | Visits to dentist or dental clinic among adults aged >= 18 years | ORH1_1 | Independent var |
| Overarching Conditions | High school completion among adults aged 18-24 years | OVC2_1 | Independent var |
| Overarching Conditions | Current lack of health insurance among adults aged 18-64 years | OVC1_1 | Independent var |
| Tobacco | Current smoking among adults aged >= 18 years | TOB1_2 | Independent var |
| Tobacco | Current smokeless tobacco use among adults aged >= 18 years | TOB2_2 | Independent var |
| Chronic Kidney Disease | Prevalence of chronic kidney disease among adults aged >= 18 years | CKD3_0 | Independent var |

```{r shape the dataaset according to the 12 applicable disease indicators}

# filter down to just the 12 selected questions
disease_with_walkability_filtered <- filter(disease_with_walkability, DataValueType == "Crude Prevalence" & 
  StratificationCategory1 == "Overall" & !is.na(NatWalkInd))
filtered_qids <- c("NPAW2_1", "NPAW11_1", "ALC5_1", "ART1_1", "AST1_1", "COPD2_0", "ORH1_1", "OVC2_1", "OVC1_1", "TOB1_2", "TOB2_2", "CKD3_0")
disease_with_walkability_filtered <- disease_with_walkability_filtered[disease_with_walkability_filtered$QuestionID %in% filtered_qids,]

# reshape the data by grouping by all the unique properties per block group, and expanding
# columns with the result of each of the relevant 12 questions for that observation.
# for now we will also keep some descriptive variables (like LocationAbbr) 
# which may come in handy for visual data exploration
collapsed_cols = c("YearStart", "LocationAbbr", "LocationDesc", "STATEFP", "COUNTYFP", "TRACTCE", "BLKGRPCE", "GEOID", "R_PCTLOWWAGE", "NatWalkInd")

disease_with_walkability_collapsed <- disease_with_walkability_filtered %>%
  pivot_wider(id_cols = collapsed_cols, names_from = c("QuestionID"), values_from = DataValueAlt) 

# rename the question columns to something easier to read
disease_with_walkability_collapsed <- disease_with_walkability_collapsed %>% 
  rename("Overweight" = "NPAW2_1") %>%
  rename("Active" = "NPAW11_1") %>%
  rename("Alcohol" = "ALC5_1") %>%
  rename("Arthritis" = "ART1_1") %>%
  rename("Asthma" = "AST1_1") %>%
  rename("COPD" = "COPD2_0") %>%
  rename("Dentist" = "ORH1_1") %>%
  rename("HighSchool" = "OVC2_1") %>%
  rename("NoInsurance" = "OVC1_1") %>%
  rename("Smoking" = "TOB1_2") %>%
  rename("Vapeing" = "TOB2_2") %>%
  rename("KidneyDisease" = "CKD3_0")
```

Next we will convert our treatment variable, NatWalkInd to binary. Our strategy will be to select the bottom 40% least walkable block groups and
consider these "not walkable" and the top 40% most walkable block groups will be considered "walkable." The middle 20% will not be used

```{r convert treatment variable to binary}
nwi40 <- quantile(disease_with_walkability_collapsed$NatWalkInd, .40) 
nwi60 <- quantile(disease_with_walkability_collapsed$NatWalkInd, .60) 
disease_with_walkability_collapsed = filter(disease_with_walkability_collapsed, NatWalkInd < nwi40 | NatWalkInd > nwi60)
disease_with_walkability_collapsed$Walkable <- ifelse(disease_with_walkability_collapsed$NatWalkInd >= nwi60, 0, 1)
# rename the dataframe to a shorter name
dww <- disease_with_walkability_collapsed
```

## Understand the variables in the dataset

The table below describes the variables that are used in this analysis, including the variables that contain the values of the questions we selected.
The "Usage In This Analysis" column categorizes how these will be used in analysis. 

| Variable Name | Variable Description | Usage In This Analysis | 
| -------- | ------------------ | -------- |
| Walkable | Binary variable, 1 if the block is walkable, 0 otherwise | Treatment Variable |
| Overweight | Percentage of the population overweight or obese | Dependent Variable |
| Active | Percentage of the population that meets physical activity guidelines | Dependent Variable |
| R_PCTLOWWAGE | Percentage of the population that makes less than $1250/month | Independent Variable |
| HighSchool | Percentage of the population ages 18-24 who have completed high school | Independent Variable |
| Alcohol | Percentage of the population with high alcohol use | Independent Variable |
| Arthritis | Percentage of the population with arthritis | Independent Variable |
| Asthma | Percentage of the population with asthma | Independent Variable |
| KidneyDisease | Percentage of the population with chronic kidney disease | Independent Variable |
| Vapeing | Percentage of the population using smokeless tobacco | Independent Variable |
| Smoking | Percentage of the population that smokes | Independent Variable |
| COPD | Percentage of the population with Chronic Obstructive Pulmonary Disease | Independent Variable |
| NoInsurance | Percentage of the population without health care coverage | Independent Variable |
| Dentist | Percentage of the population with dentist visits | Independent Variable |

## Get familiar with the data using summary statistics

The first thing we will note is that what started as a very large data set with almost a million observations has diminished down to a relatively small sample size (318 observations) with all of the necessary filtering and aggregation that was done. We will need to do the best we can with it. Due to the small sample size, propensity score matching may not be a viable option, and generally we need to take care with the conclusions we draw from the analysis. 

```{r Look at the structure of the data.frame}
str(dww)
```
Another observation from the summary statistics is that there are a lot of NA values for the disease indicators (such as Alcohol, Arthritis, Asthma, etc. ). This will also impact the confidence of any estimated causality conclusions drawn from this data. 

```{r summary statistics}
summary(dww)
```
Here we take a look at the correlation between the variables. The rendering in markdown is a bit small to see so you can also un-comment the "ggsave" command to export a larger rendering. Key observations:
1. Walkable (the treatment variable) is correlated with higher levels of alcohol use, smoking, vapeing, and noinsurance. 
2. Overweight (the dependent variable) is correlated with lower likeiyhood of high school graduation, lower physical activity, and dental visits, and higher levels of arthritis, COPD, smoking and vapeing.

```{r pairs summary of the data, message=FALSE, warning=FALSE}
dww_numeric = dww %>% select(-YearStart, -LocationAbbr, -LocationDesc, -STATEFP, -COUNTYFP, 
       -TRACTCE, -BLKGRPCE, -GEOID)
#pairs(dww_numeric)
dww_ggpairs = GGally::ggpairs(dww_numeric)
#ggsave("dww_ggpairs.png", plot=dww_ggpairs, width=20, height=20)
dww_ggpairs
```


## TODO: Insert rest of paper here

## Appendix

### Original data pre-processing methodology

As described in the objective section, the original data came from two sources. The disease indicators data contains location information in the form of latitude and longitude. The walkability data contains location information in the form of Federal census location codes (FIPS codes). The pre-processing technique below was used to convert the latitude and longitude to FIPS codes, and then perform a join operation utilizing the FIPS codes. The resulting data is the original disease indicators data, augmented with the walkability information for the location corresponding to the original latitude and longitude. 

In other words, for every row in the disease indicators data set, the corresponding walkability information for the region was added to that row.
All of the commands are commented out to prevent them from being executed on knit since they take a long time to run.

#### Download the raw data

```{r Download the raw data, echo=TRUE }
#download.file("https://edg.epa.gov/EPADataCommons/public/OA/EPA_SmartLocationDatabase_V3_Jan_2021_Final.csv", destfile="walkability.csv")
#download.file("https://data.cdc.gov/api/views/g4ie-h725/rows.csv?accessType=DOWNLOAD", destfile="diseaseindicators.csv")
```

#### Load the data into R

```{r load the raw data, echo=TRUE}
#walkability <- read.csv("walkability.csv")
## some of the disease data has no GeoLocation, which we cannot use for our analysis, so filter those out
#disease <- filter(read.csv("diseaseindicators.csv"), GeoLocation != "")
```

#### Extract the latitude and longitude into separate columns

```{r split disease long and lat to separate columns}

## Extract the latitude and longitude values from the GeoLocation column using str_extract_all()
#geo_df <- str_extract_all(disease$GeoLocation, "-?[0-9]+\\.[0-9]+")

## Convert the extracted values to numeric and assign them to the corresponding latitude and longitude columns
#disease$lat <- as.numeric(sapply(geo_df, function(x) x[2]))
#disease$long <- as.numeric(sapply(geo_df, function(x) x[1]))

```

#### Fetch the geographic information required to map latitude and logitude to FIPS blocks

The tigris library provides a function "block_groups" which returns geographic information about every FIPS block. This geographic information can be used to convert latitude and longitude to FIPS block. The following code downloads all of the block_groups for every block in the walkability data set.

``` {r fetch geographies for blocks}

## create data frame for block_groups data
#allblockgroups <- data.frame(matrix(ncol=6, nrow=0))
#colnames(allblockgroups) <- c('STATEFP', 'COUNTYFP', 'TRACTCE', 'BLKGRPCE', 'GEOID', 'geometry')

## get block geography data for each state in the walkability dataset
#stateCodes <- data.frame(unique(walkability$STATEFP))
#for (i in 1:nrow(stateCodes)) {
#  stateCode=stateCodes[[1]][i]
#  counties = distinct(filter(walkability, STATEFP == stateCode), COUNTYFP)$COUNTYFP
#  new_blocks <- block_groups(state=stateCodes[[1]][i], counties) %>%
#    select(STATEFP, COUNTYFP, TRACTCE, BLKGRPCE, GEOID, geometry)
#  allblockgroups <- rbind(allblockgroups, new_blocks)
#}

```

#### Use block geographies to convert longitude and latitude to FIPS blocks

``` {r convert disease long and lat to geoid10 }

#my_points <- data.frame(
#  x = disease$lat,
#  y = disease$long
#) %>%
#  st_as_sf(coords = c("y", "x"),
#     crs = st_crs(allblockgroups))
  
#my_points_blocks <- st_join(my_points, allblockgroups)
#disease$STATEFP = as.integer(my_points_blocks$STATEFP)
#disease$COUNTYFP = as.integer(my_points_blocks$COUNTYFP)
#disease$TRACTCE = as.integer(my_points_blocks$TRACTCE)
#disease$BLKGRPCE = as.integer(my_points_blocks$BLKGRPCE)
#disease$GEOID = as.numeric(my_points_blocks$GEOID)

```

#### Join the disease indicators and walkability data sets based on FIPS blocks

``` {r join the disease and walkability data }

# Join the disease data with the walkability data
#disease_with_walkability <- left_join(disease, walkability, 
#                                  by = c("STATEFP", "COUNTYFP", "TRACTCE", "BLKGRPCE"))

```

#### Export the joined data to be used for further processing later.

``` {r export the processed and merged data to a csv}
#write.csv(disease_with_walkability, file = "disease_with_walkability.csv")
```
